\chapter{Introduction}\label{ch:introduction}

In this thesis, we address a long-standing challenge in modern experimental nuclear physics that stems from the difficulty of describing the breadth of reactions that occur in a given experiment. The efficiency needed from detectors used in the analysis of rare nuclides means that much of the collected data is not pertinent to the questions we ask. One way of filtering out these unwanted events is to use algorithms from machine learning; which is a field of study with elements from mathematics and computer science that deals with pattern recognition and function approximation. In this thesis, we explore the application of machine learning to the segmentation of events from a rare isotope detector. In this way, we address the challenging aspect of the experiment by not carrying \textit{a priori} assumptions about the reactions that occur. 

Nuclear physics is the pursuit of understanding nuclides, which are the building blocks of the visible universe. These building blocks are made up of protons and neutrons. However, while most matter we interact with is stable, the vast majority of the hitherto discovered nuclides are not. Understanding these unstable nuclides are essential to our understanding of the universe, as well as having implications in medicine and the industry. Detecting these rare nuclides require very sensitive and specialized equipment. One such piece of equipment is the active target time-projection chamber (AT-TPC) detector. When running, it can detect on the order of $10^4$ events each hour, producing terabytes of data each day it is active. In this thesis, we work with data from an AT-TPC detector constructed and commissioned at the National Cyclotron Laboratory located on the Michigan State University campus. 

The two main topics we address in this thesis is how many events from known reactions do we need to construct a good model, and given the absence of known reactions can we segment events by reaction type. 

\section{Machine learning}

In modern science, data analysis has become ubiquitous. For many applications, it is now possible to collect data with samples numbering from several thousand to billions, which has transformed the modeling needs in those sciences. Machine learning is closely tied to this development. Machine learning models tend to have a large number of degrees of freedom, which means they need large volumes of data to perform well. These models have for example been used to beat human professionals in chess, and go\footnote{Go is a strategy game played on a grid, and is like chess in that it rewards long-term planning.} from just knowing the rules of the game \cite{Silver2018}. They also fundamentally affect our on-line interactions\footnote{\href{https://research.fb.com/category/machine-learning/}{Facebook} invests heavily in machine learning research, and apply insights from this research to dictate what we view and interact with on their platforms.}. However, machine learning models also include simpler methods, like ordinary least squares regression. 

At the heart of modern machine learning is the Artificial Neural Network family of algorithms. These models, which are based on systems of biological neurons, have shown themselves to be very good at approximating complex functions \cite{Lin2017}. They are also fundamentally flexible algorithms and can are applied to image analysis, time-series prediction, or regression tasks with relative ease.

In this thesis, we will consider a particular property of Neural Networks. We know that they can achieve a high degree of compression of complex input and that these compressions can inform us of salient differences in the input space. We consider the space of nuclear reactions in an experiment in this thesis and apply methods that can discover structure without being explicitly informed of what to look for.

\section{Why machine learning?}

Traditionally, data from experiments recorded with an AT-TPC is analyzed with a Monte Carlo method. In this framework, each event is treated as a potential candidate for the reaction of interest. Subsequently, several physical parameters are tested to find the best fit for that event. Once all events in the dataset are fit, a threshold value for the fit-statistic is chosen. Events that are below this threshold value are then said to be events of interest, and the rest are discarded. Some of the events of interest are bound to be discarded with this method, which is problematic when the interesting reactions are rare. 

In addition to these challenges, this method of analysis turns out to be very computationally costly. Furthermore, the presumption that each event is considered a possible candidate for the reaction of interest can be problematic. In the case where the breadth of possible reactions is not known, fitting against the reaction of interest can give unexpected results. Lastly, the fitting method requires that the records are of complete tracks, but this is not always the case.

In this thesis, we then propose computationally feasible models, that do not explicitly fit against the reaction of interest, and who are agnostic to the completeness of the tracks, from the machine learning literature.

\section{Ethical considerations}

Reproducibility is a cornerstone of the scientific process, and in computational science version control systems like \href{https://GitHub.com/}{GitHub} make it possible to track development. Additionally, version control provides opportunities for other researchers to reproduce results easily. To further this standard of research we make code from this thesis available for all to access from our \href{https://GitHub.com/ATTPC/VAE-event-classification}{repository.} 

In addition to being applied in quantum and statistical mechanics, and the design of robots with robust movement abilities\footnote{Boston dynamics design and manufacture \href{https://www.youtube.com/watch?v=LikxFZZO2sk}{robots} with exceptional adaptability to varying environments}, machine learning has also been applied to the detection of sexual orientation from facial images \cite{Wang2018}. The latter gives rise to serious ethical concerns as LGBTQ persons are still persecuted in many nations, algorithms like these then have the potential to influence these persons livelihood, as well as their mental and physical health. The algorithms developed for this thesis are latent variable models for the segmentation of data, and it is easy to transfer these algorithms to different arenas e.g., the segmentation of people in discrete groups.

It is difficult to say what moral obligations the researcher carries in this instance. However, this is not unfamiliar territory, as the development of nuclear weaponry is closely tied with the emergence of modern nuclear theory. From this history, we can infer that the openness of the scientific discourse is a necessary ingredient in providing strategies for dealing with complex issues such as these. That is not to say that this discourse is sufficient.  Being critical of our role as researchers in the development of these algorithms is essential Additionally, we must engage with society and lawmakers on these issues, as part of the responsibility of being a scientist. 

\section{Structure of the thesis}

As machine learning is still nascent in its application to nuclear physics, we begin with a thorough introduction to the theory in the first chapter of part I. We continue in the second chapter with an introduction to the models implemented and applied in this thesis. The last chapter of part I is devoted to elaborating on the experiment, and data, that forms the basis for our analysis.

Part II is devoted to the details of the implementation. We have chosen the \lstinline{Python} programming language as the basis for the implementation because of the mature machine learning libraries developed for the language. 

Lastly, in parts III and IV, we present our findings and discuss those. We also present avenues for further research based on our findings.
